Supporting Information

### 2015-12-18 ms.Rmd
Initial submission TPJ

| ^§^ Institut de Recherche pour le Développement, 911 Avenue Agropolis, 34394 Montpellier, France.
| ^¶^ Università degli Studi di Milano, Via Celoria 26, 20133 Milano, Italy.

## Running title

Expression profile of rice inflorescence meristems

## Keywords

Inflorescence development, rice, reproductive meristems, RNA sequencing

## **Sequencing library repository**

## Word count

@@

Understanding the events that occur during development that determine panicle characteristics, such as branching complexity and its plasticity, will be vital for sustainable improvement of rice yield potential using targeted breeding programs. Despite advances in the characterization of individual genes and their interactions, a complete understanding of the control of panicle morphology and grain yield will require mechanistic studies explaining the interactions of gene regulatory network components with each other and with the environment [@Azpeitia:2013jia]. One step towards this goal is to describe the differences in gene expression between different meristem types during development. In this article, we describe the measurement and analysis of genome-wide expression in meristematic tissues from the early stages of panicle development in *O. sativa* ssp. *japonica* cv. Nipponbare, using a precise laser microdissection and RNA sequencing approach.

@@

Significant changes in gene expression between apical and axillary meristems are supported by two observations: a difference in expression of annotated transcription factor genes between RM and PBM samples, and the recovery of co-expression clusters of primarily uncharacterized genes that appear to be either switched on or off between the RM and the PBM. Clusters `r getClusterName('dec', 'small')` and `r getClusterName('inc', 'small')` contain genes that change in expression between apical (RM) and axillary meristems (PBM–SM). Axillary meristem initiation requires auxin synthesis and transport in *A. thaliana* and maize [reviewed by @Gallavotti:2013ki], but these processes are not as well understood in rice. Cluster `r getClusterName('inc', 'small')` contains the auxin response factor *ARF6A* and the small auxin-up RNA *SAUR33*, as well as two ABC transporters, while the complementary cluster `r getClusterName('dec', 'small')` contains four uncharacterized genes with annotations relating to auxin (*IAA10* and *IAA16*, two auxin-responsive transcriptional regulators; *ABCB27*, an ABC transporter; and *LOC_Os01g63770*, an AUX/LAX transporter), and the GH3 gene related to auxin homeostasis, *GH3-8*. In *A. thaliana*, excess auxin may be inactivated by conjugation to amino acids by GH3 enzymes [@Staswick:2005cv], and transgenic rice plants containing 3 kb of upstream sequence from *FLO-LFY HOMOLOG OF RICE* (*RFL*) fused to the *GH3-8* coding sequence, presumably expressing *GH3-8* in the endogenous pattern of *RFL* [*i.e.* PBM and SBM; @IkedaKawakatsu:2012co], have smaller panicles with fewer branches [@Yadav:2011db]. GH3-8 enzyme conjugates amino acids to auxin *in vitro* [@Ding:2008hv], suggesting that the panicle phenotype may be related to reduced auxin signalling. Although auxin response factors and Aux/IAA response genes are generally more strongly expressed at earlier stages (`r I(pasteLabel("Figure ", figCount, "f_gsea"))`), the presence of auxin-related genes in complementary clusters suggests that the transcriptional effects of auxin signalling are dependent on meristem type, and understanding of the relationship between auxin and meristem specification will require functional analysis of the components of the regulatory network involved in the response in these tissues.

@@

*Oryza sativa* ssp. *japonica* cv. Nipponbare plants were grown in a growth chamber with 70% relative humidity at 30°C during the day and 26°C at night. After 7 weeks in long day conditions (14 hours light and 10 hours darkness) plants were transferred to short day conditions (10 hours light and 14 hours darkness) to induce floral transition. Inflorescence meristems were harvested from 6 to 14 days after the change of photoperiod.

@@

**`r I(pasteLabel("Figure S", s_figCount, "sf_madsTree"))`**: Differential expression
of genes encoding putative MIKC proteins in *A. thaliana*, *Solanum lycopersicum* and *Oryza sativa* ssp. *japonica*.

>>>>

| ^§^ Institut de Recherche pour le Développement, UMR DIADE, 911 Avenue Agropolis, 34394 Montpellier, France.
| ^¶^ Università degli Studi di Milano, Via Celoria 26, 20133 Milano, Italy.

**Running title**: Expression profile of rice inflorescence meristems

**Keywords**: Inflorescence development, rice, reproductive meristems, RNA sequencing

**Data availability**: Unprocessed sequencing libraries have been uploaded to the NCBI Sequence Read Archive under accession SRP067488.

**Word count**: 6,242 words.

@@

Understanding the events that occur during development that determine panicle characteristics, such as branching complexity and its plasticity, will be vital for sustainable improvement of rice yield potential using targeted breeding programs. Despite advances in the characterization of individual genes and their interactions, a complete understanding of the control of panicle morphology and grain yield will require mechanistic studies explaining the interactions of gene regulatory network components with each other and with the environment [@Azpeitia:2013jia]. One step towards this goal is to describe the differences in gene expression between different meristem types during development. In this article, we describe the measurement and analysis of genome-wide expression in meristematic tissues from the early stages of panicle development in *Oryza sativa* ssp. *japonica* cv. Nipponbare, using a precise laser microdissection and RNA sequencing approach.

@@

Significant changes in gene expression between apical and axillary meristems are supported by two observations: a difference in expression of annotated transcription factor genes between RM and PBM samples, and the recovery of co-expression clusters of primarily uncharacterized genes that appear to be either switched on or off between the RM and the PBM. Clusters `r getClusterName('dec', 'small')` and `r getClusterName('inc', 'small')` contain genes that change in expression between apical (RM) and axillary meristems (PBM–SM). Axillary meristem initiation requires auxin synthesis and transport in *Arabidopsis thaliana* and maize [reviewed by @Gallavotti:2013ki], but these processes are not as well understood in rice. Cluster `r getClusterName('inc', 'small')` contains the auxin response factor *ARF6A* and the small auxin-up RNA *SAUR33*, as well as two ABC transporters, while the complementary cluster `r getClusterName('dec', 'small')` contains four uncharacterized genes with annotations relating to auxin (*IAA10* and *IAA16*, two auxin-responsive transcriptional regulators; *ABCB27*, an ABC transporter; and *LOC_Os01g63770*, an AUX/LAX transporter), and the GH3 gene related to auxin homeostasis, *GH3-8*. In *A. thaliana*, excess auxin may be inactivated by conjugation to amino acids by GH3 enzymes [@Staswick:2005cv], and transgenic rice plants containing 3 kb of upstream sequence from *FLO-LFY HOMOLOG OF RICE* (*RFL*) fused to the *GH3-8* coding sequence, presumably expressing *GH3-8* in the endogenous pattern of *RFL* [*i.e.* PBM and SBM; @IkedaKawakatsu:2012co], have smaller panicles with fewer branches [@Yadav:2011db]. GH3-8 enzyme conjugates amino acids to auxin *in vitro* [@Ding:2008hv], suggesting that the panicle phenotype may be related to reduced auxin signalling. Although auxin response factors and Aux/IAA response genes are generally more strongly expressed at earlier stages (`r I(pasteLabel("Figure ", figCount, "f_gsea"))`), the presence of auxin-related genes in complementary clusters suggests that the transcriptional effects of auxin signalling are dependent on meristem type, and understanding of the relationship between auxin and meristem specification will require functional analysis of the components of the regulatory network involved in the response in these tissues.

@@

*O. sativa* ssp. *japonica* cv. Nipponbare plants were grown in a growth chamber with 70% relative humidity at 30°C during the day and 26°C at night. After 7 weeks in long day conditions (14 hours light and 10 hours darkness) plants were transferred to short day conditions (10 hours light and 14 hours darkness) to induce floral transition. Inflorescence meristems were harvested from 6 to 14 days after the change of photoperiod.

@@

**`r I(pasteLabel("Figure S", s_figCount, "sf_madsTree"))`**: Differential expression
of genes encoding putative MIKC proteins in *A. thaliana*, *S. lycopersicum* and *O. sativa* ssp. *japonica*.

### 2015-12-18 si.Rmd
Initial submission TPJ

@@

# remove the first hline
st_libStats_x_print <- sub("\\begin{tabular}{rrrrrrrrrrrrrrrr}\n  \\hline\n", "\\begin{tabular}{rrrrrrrrrrrrrrrr}\n", st_libStats_x_print, fixed = TRUE)
cat(st_libStats_x_print)
cat("¹ ePBM/AM")
```

@@

```{r sf_madsTree, fig.height = 9.535, fig.cap = paste0("Differential expression of genes encoding putative MIKC proteins in \\emph{A. thaliana}, \\emph{Solanum lycopersicum} and \\emph{Oryza sativa} ssp. \\emph{japonica}. The tree is a neighbour-joining tree estimated from genetic distance of amino acid sequences (see Methods S1). Named proteins starting with AT\\_ or OS\\_ are from Arabidopsis or rice respectively, with synonyms in parentheses, and other proteins are labelled with their unique identifiers.", "\\newline ", "The colour of the label represents the Log\\textsubscript{2}-fold change (L\\textsubscript{2}FC) in expression of the gene encoding that protein between undifferentiated and differentiated meristems. Where there is no coloured label, the adjusted \\emph{p}-value associated with the difference in expression was greater than 0.1, or the gene for the protein in question was excluded from the differential expression analysis by \\texttt{DESeq2}, which may occur because of low read counts, an extreme count outlier or the gene failing independent filtering (Love \\emph{et al}., 2014).", "\\newline ", "For rice, L\\textsubscript{2}FC was estimated from the RM and SM samples reported here. For \\emph{A. thaliana}, L\\textsubscript{2}FC was estimated using the inflorescence meristem and floret meristem samples reported by Mantegazza \\emph{et al}. (2014). The sympodial inflorescence meristem and floret meristem samples generated by Park \\emph{et al}. (2012) were used to estimate L\\textsubscript{2}FC values for tomato genes.")}
print(sf_madsTree)
```

\clearpage

# Methods S1

## Comparative analysis of the MADS family

The primary sequences of *Arabidopsis thaliana*, *Solanum lycopersicum* and *Oryza sativa* MADS proteins were downloaded by querying Phytozome 10.3 [@Goodstein:2012ho] for the PANTHER family PTHR11945 ('MADS BOX PROTEIN') via the `biomaRt` interface [@Thomas:2003db; @Durinck:2009ki]. An initial alignment of all proteins was performed with `Clustal Omega` using a full distance matrix [@Sievers:2011fn]. To clean the alignment, poorly conserved regions where greater than `r minpcnongap` percent of sequences had gaps or less than `r minpcident` percent of pairwise comparisons between sequences had identical residues were removed. To recover MADS genes related to the Arabidopsis MIKC clade, a UGPMA tree was estimated from the cleaned alignment and the tree was cut at the lowest height that returned a clade including all type II Arabidopsis MADS proteins listed in the Arabidopsis transcription factor database on the Arabidopsis Gene Regulatory Information Server [@Yilmaz:2011ew], with as few type I proteins as possible. Protein sequences in this clade that were longer than `r minProtLength` amino acids were realigned using `Clustal Omega`, and the alignment was cleaned as described above and used to estimate a neighbour-joining tree with the `bionj` algorithm implemented in the `ape` package [@Paradis:2004dv] rooted to type I MADS sequences (`r og`). L~2~FC change expression values for tomato and Arabidopsis genes were estimated from samples that have previously been described [@Mantegazza:2014di; @Park:2012dg] using the same `STAR`–`DESeq2` pipeline described in the main text for the rice samples [@Dobin:2012fg; @Love:2014do]. The tree was plotted using the `ggtree` extension for `ggplot2` [https://github.com/GuangchuangYu/ggtree; @Wickham:2009tn], with clade labels following the nomenclature presented by [@Arora:2007fa].

\clearpage

>>>>

```{r sf_madsTree, fig.height = 9.535, fig.cap = paste0("Differential expression of genes encoding putative MIKC proteins in \\emph{A. thaliana}, \\emph{S. lycopersicum} and \\emph{O. sativa} ssp. \\emph{japonica}. The tree is a neighbour-joining tree estimated from genetic distance of amino acid sequences (see Methods S1). Named proteins starting with AT\\_ or OS\\_ are from Arabidopsis or rice respectively, with synonyms in parentheses, and other proteins are labelled with their unique identifiers.", "\\newline ", "The colour of the label represents the Log\\textsubscript{2}-fold change (L\\textsubscript{2}FC) in expression of the gene encoding that protein between undifferentiated and differentiated meristems. Where there is no coloured label, the adjusted \\emph{p}-value associated with the difference in expression was greater than 0.1, or the gene for the protein in question was excluded from the differential expression analysis by \\texttt{DESeq2}, which may occur because of low read counts, an extreme count outlier or the gene failing independent filtering (Love \\emph{et al}., 2014).", "\\newline ", "For rice, L\\textsubscript{2}FC was estimated from the RM and SM samples reported here. For \\emph{A. thaliana}, L\\textsubscript{2}FC was estimated using the inflorescence meristem and floret meristem samples reported by Mantegazza \\emph{et al}. (2014). The sympodial inflorescence meristem and floret meristem samples generated by Park \\emph{et al}. (2012) were used to estimate L\\textsubscript{2}FC values for tomato genes.")}
print(sf_madsTree)
```

@@

# remove the first hline
st_libStats_x_print <- sub("\\begin{tabular}{rrrrrrrrrrrrrrrrr}\n  \\hline\n", "\\begin{tabular}{rrrrrrrrrrrrrrrrr}\n", st_libStats_x_print, fixed = TRUE)
cat(st_libStats_x_print)
cat("¹ ePBM/AM")
```

@@

# Methods S1: Comparative analysis of the MADS family

The primary sequences of *A. thaliana*, *S. lycopersicum* and *O. sativa* MADS proteins were downloaded by querying Phytozome 10.3 [@Goodstein:2012ho] for the PANTHER family PTHR11945 ('MADS BOX PROTEIN') via the `biomaRt` interface [@Thomas:2003db; @Durinck:2009ki]. An initial alignment of all proteins was performed with `Clustal Omega` using a full distance matrix [@Sievers:2011fn]. To clean the alignment, poorly conserved regions where greater than `r minpcnongap` percent of sequences had gaps or less than `r minpcident` percent of pairwise comparisons between sequences had identical residues were removed. To recover MADS genes related to the Arabidopsis MIKC clade, a UGPMA tree was estimated from the cleaned alignment and the tree was cut at the lowest height that returned a clade including all type II Arabidopsis MADS proteins listed in the Arabidopsis transcription factor database on the Arabidopsis Gene Regulatory Information Server [@Yilmaz:2011ew], with as few type I proteins as possible. Protein sequences in this clade that were longer than `r minProtLength` amino acids were realigned using `Clustal Omega`, and the alignment was cleaned as described above and used to estimate a neighbour-joining tree with the `bionj` algorithm implemented in the `ape` package [@Paradis:2004dv] rooted to type I MADS sequences (`r og`). L~2~FC change expression values for tomato and Arabidopsis genes were estimated from samples that have previously been described [@Mantegazza:2014di; @Park:2012dg] using the same `STAR`–`DESeq2` pipeline described in the main text for the rice samples [@Dobin:2012fg; @Love:2014do]. The tree was plotted using the `ggtree` extension for `ggplot2` [https://github.com/GuangchuangYu/ggtree; @Wickham:2009tn], with clade labels following the nomenclature presented by @Arora:2007fa.
